### 2015-08-24 manuscript/content/method.tex
Intro. edits
Paper additions and cleanup.

\section{Methodology}
Research involved people to volunteer in a study that would take roughly two hours in their own home with no monetary compensation, but alcoholic drinks provided free of charge. During the study, volunteers would be measured (height in inches) and weighed (weight in pounds and converted to Kg). This information was recorded and used to determine the amount of alcohol required for the user to reach a BAC of .08 in 1.5 hours by the following formula:

\begin{figure}[H]
	\centering
	\includegraphics[scale=.40]{formula.png}
	\caption{EBAC Formula}
\end{figure}

The volunteer was then given the Microsoft Band to wear on the wrist of their dominant hand and data collection began. Over the course of two hours, the volunteer participated in simple card games while drinking a predetermined number of 1.5oz. of 60 proof liquor shots. While the volunteer took his drink, a researcher would hold the button on the application in order to label and record the event. Also, right before the second shot and every 25 minutes following, the volunteers BAC was measured with the use of a breathalyzer and then recorded in the application. After 1.5 hours, the volunteer ceased drinking but data collection from the smartwatch and breathalyzer continued. At the end of the two hours, BAC was measured one final time and recorded. Data then collection ended and the Microsoft Band was removed. The volunteer was advised to remain safe within their home until sober.

\begin{figure}[H]
	\centering
	\includegraphics[scale=.28]{datacollectionprocess.png}
	\caption{Data collection}
\end{figure}

Collected data was uploaded to a web server and then stored on a mySQL database to be used for visualization of data on a web client. Data was also reviewed for clean up and to find any correlations between the sensor data and measure BAC. This data was then used to train a machine learning model.

>>>>

\section{Methodology}

In this section, we will describe how we collected the data, we will present an analysis of the data, and then lead into the discussion of the machine learning models used.

\subsection{Collection}

Our collection began with the development of an Android application that connected to a Microsoft Band smartwatch and collected data from all of its available sensors. This development led to the general Android-based sensor gateway discussed in the previous section. Next, we developed a general procedure for our volunteers to follow during the collection of data. Our volunteers were eager to freely contribute the anonymous data used in this paper. The system collected the data into a .csv file on the Android smartphone and was also transmitted to a central MySQL server.

The Android platform was version 5.0 (kernel 3.4.0-4432708) running on a Samsung S5 smartphone. The Microsoft Band was Build Version 10.2.2818.0 09 R. Samples of the sensor data were collected every three seconds, based on the update speed of the Band's heart rate sensor. For every sample, the most recent sensor value was used for every sensor if available, else it would use the last updated value; or in the case of the accelerometer and gyroscope sensors, the last three values were averaged with linear weighting (the most recent having the most importance). There may be a better weighting, but this weighting was suitable for our purposes.

We designed a simple, two-hour procedure for the collection of the data. First, we established some necessary information about the subject to estimate the amount of alcohol necessary to reach 0.08 BAC in a 1.5 hour period using the Widmark equation (1). The particular formulation of it we used is the following: \begin{equation}
SD =  BW \cdot Wt \cdot (EBAC + (MR \cdot DP)) \cdot 0.4690
\end{equation} where $SD$ is the number of standard drinks (10 grams ethanol), $BW$ is the body water contant (0.58 for men and 0.49 for women), Wt is the body weight in lbs, EBAC is the estimated BAC, MR is the metabolism rate (0.17 for women and 0.18 for men), DP is the drinking period in hours, and  $0.4690 = 0.4536 \div (0.806 \cdot 1.2)$, a combination of two constants from the equation and a converstion from kg to lbs \cite{Andersson:2009}\cite{Wiki:BAC}. This amount was used to estimate the number of standard drinks to be consumed over the set time period, distributed over equal intervals. During this process, at every 25 minutes we took a measurement of the BAC using a BACtrack Trace\textsuperscript{TM} Pro breathalyzer. This measurement interval was determined by the cooldown rate of the breathalyzer. The activity chosen for the volunteers to engage in was a card or board game of their choice. Drinking stopped before 1.5 hours while collection and BAC measurement continued for another 30-45 minutes.

\subsection{Data Analysis}

\subsection{Feature Selection}

\subsection{Model Building}

### 2015-08-24 manuscript/content/related.tex
Paper additions and cleanup.

\section{Related Work}

There are a few ways of approaching the problem of determining a person's blood alcohol content. One approach is to model mathematically the elimination of ethanol in the human body. In this case, the Widmark equation, published in 1932 by E.M.P. Widmark, is a very popular one, \begin{equation}
•C=\frac{A}{rW}-(\beta t)
\end{equation} where C is the BAC, r and $\beta$ are empirically determined constants, A is the mass of the consumed alcohol, and W is the body weight of the person.
	These days, there have been several improvements and variants. Douglas Posey and Ashraf Mozayani published an excellent article comparing this model using parameters determined by different researchers and discussing different models \cite{Posey:2007}. They found that the Widmark equation tends to overestimate, and that there can be significant discrepancy between the results of the different models. Despite that, they do provide a rough estimate. The problem is that these models also require a good deal of information that prohibit their use in a non-intrusive, drunkenness warning system.

Another approach to the problem is simply to directly measure the BAC, such as with the recent investigations into the use of transdermal ethanol sensors. Transdermal sensing provides a more discreet way to measure BAC, but this method is accompanied by the problem of a significant time lag between the sensed alcohol concentration and actual blood alcohol concentration. Gregory Webster and Hampton Gabler closely investigated this problem. They found that the lag is predictable, but variable, and requires additional information about the drinks taken to accurately predict \cite{Webster:2007}.

In 2014, James Baldwin applied for a patent on a wearable blood alcohol measuring device \cite{Baldwin:2014} as a non-invasive, discreet, and wearable method of continuously predicting blood alcohol content. Baldwin's method involves the use of a transdermal ethanol sensor and a method of predicting future BAC given the user's intended consumption. It is also meant to provide guidance for the user as to when to stop in order to be at a certain BAC by a previously specified time. Baldwin's invention is very similar to this project with the only major difference being that it uses a wearable transdermal ethanol sensor to directly measure BAC instead of using common sensors found in smartwatches and a prediction model, such as we do, in order to predict BAC. The use of a transdermal ethanol sensor is proven to have many issues, mostly in the delay between consumption and sensing \cite{Webster:2007}. Another issue is that it requires a special sensor that would only be used for this purpose. Our project improves on this issue by using a common device that many already own for various other reasons. Also, due to our approach for monitoring drunkenness, our method may provide more immediate predictions.

In 2015, Koukiou and Anastassopoulos published a paper on using a neural network to identify drunkenness from thermal infrared imagery \cite{Koukiou:2015}. Different neural networks were trained on different parts of the face in order to determine which areas can be used to discriminated "a drunk from a sober person." Their findings showed that the forehead was the main feature to change thermal behavior from the consumption of alcohol. Later, the whole face was used to train a neural structure which showed high performance on discrimination. The results showed that the forehead and the nose are the areas of the face that were most prominent in drunk people. A neural networks was then trained on the forehead of only one person which was then used on other persons in order to determine those who were drunk. The results had a success rate of around 90\%. This study shows that there is a correlation between skin temperature and drunkenness, which is applicable to our research because of the skin temperature sensor found on many smartwatches. This project has many similar applications as our would, such as being able to be incorporated into ignition interlock systems or to give evidence to authorities about intoxication in investigations. However, this system may not be able to be used by people to determine their own drunkenness as it requires a thermal infrared camera to be focused on the users face in order to do so. This would be difficult to do discreetly or conveniently.

There is also research going into using various sensors for health applications, many using a machine learning approach to do so.

In "Multi-Sensor Fusion for Enhanced Contextual Awareness of Everyday Activites with Ubiquitous Devices" \cite{Guiry:2014}, the authors attempt to use the sensors in smartphones and smartwatches to identify various daily activities such as walking, running, cycling, sitting, etc. It also identified whether or not the user is indoors or outdoors. Collected data was then used to train 5 different machine learning algorithms: C4.5, CART, Na\"{\i}ve Bayes, Multi-Layer Perceptrons, and Support Vector Machines. Results showed promise, with one model for inside/outside classification being 100\% correct. This study shows that the use of multiple sensors improves classification accuracy. For our research, we may be able to improve BAC prediction by recognizing when the user is taking a drink. This research shows that machine learning models can be used to label collected sensor data to certain activities.

Study has been done in using accelerometers to detect seizures \cite{Milosevic:2014} and using smartphones and smartwatches to monitor Parkinson Disease \cite{Sharma:2014}. In research on the detection of seizures \cite{Milosevic:2014}, researches use accelerometers on each wrist and ankle and a machine learning model in order to detect epileptic convulsions. Results included detection of 19 out of 23 seizures that were longer than 30 seconds each, with only a false alarm rate of .39 per hour. SPARK, a proposed framework that uses smartphones and smartwatches to monitor for symptoms of Parkinson Disease \cite{Sharma:2014}, including speech analysis, facial tremors, dyskinesia, freezing of gait. SPARK is for users to have their body motions monitored for dyskinesia and freezing gait while also having the user read paragraphs for speech analysis or pointing the camera for facial analysis. SPARK will also provide recommendations to physicians based on the collected information, such as changes in medication. SPARK is still being tested in order to determine the usefulness of such a framework for those who would use it. It is noted that there are limitations to it in that many patients will not be able to adopt the technologies required for it. There are also many issues, such as sensor misplacement and log errors, that can be encountered. Both these project show a desire to create health-based monitoring applications.

%Current list of applications on smart watch, and their novelty. Current Prediction of BAC and associated ML algorithms

>>>>

\section{Related Work}
There are a few ways of approaching the problem of determining a person's blood alcohol content. One approach is to model mathematically the elimination of ethanol in the human body. In this case, the Widmark equation, published in 1932 by E.M.P. Widmark, is a very popular one, \begin{equation}
•C=\frac{A}{rW}-(\beta t)
\end{equation} where $C$ is the BAC, $r$ and $\beta$ are empirically determined constants, $A$ is the mass of the consumed alcohol, and $W$ is the body weight of the person.
	These days, there have been several improvements and variants. Douglas Posey and Ashraf Mozayani published an excellent article comparing this model using parameters determined by different researchers and discussing different models \cite{Posey:2007}. They found that the Widmark equation tends to overestimate, and that there can be significant discrepancy between the results of the different models. Despite that, they do provide a rough estimate. The problem is that these models also require a good deal of information that prohibit their use in a non-intrusive, drunkenness warning system.

Another approach to the problem is simply to measure the BAC directly. Transdermal ethanol sensors have been a recent option for this approach. These can provide a discreet way to measure intoxication, but they are accompanied by the problem of a significant time lag between the sensed alcohol concentration and actual blood alcohol concentration. Gregory D. Webster and Hampton C. Gabler closely investigated this problem. They found that the lag is predictable, but not constant, and requires additional information about the drinks taken to accurately predict \cite{Webster:2007}.

Similar to our project, James A. Baldwin has a patent on a system involving a wearable transdermal ethanol sensor and a mobile device to capture the information \cite{Baldwin:2014}. Baldwin describes his system as using a mathematical model to predict the user's BAC given the transdermal sensor data and information about the drinks the user plans to consume. A benefit of our system is that it involves no input by the user about the drinks taken, and the user need not buy a special sensor dedicated to this task alone.

Aside from measuring BAC directly, or developing a biologically-based mathematical model, machine learning is another good approach. Georgia Koukiou and Vassilis Anastassopoulos published research this year in using a neural network to identify drunkenness from thermal infrared images of peoples' faces \cite{Koukiou:2015}. Neural networks were trained on different parts of the face in order to determine which areas can be used to classify drunkenness. They found the forehead was the most significant facial location to observe for determining the drunkenness of a person. Their study takes advantage of the effect of alcohol making blood vessels dilate allowing warm blood to come closer to the skin; which is also an important effect for our research. Such a system may be good for ignition interlock systems, or drunk surveillance.

Outside of BAC studies, there has been plenty of research into detecting other activities using smartphone and smartwatch sensor data. In \cite{Guiry:2014}, John J. Guiry, Pepijin van de Ven, John Nelson, attempt using the sensor data to identify various daily activities, such as: walking, running, cycling, and sitting. In their study, they use several machine learning algorithms for their approach: C4.5, CART, Na\"{\i}ve Bayes, ANN, and SVM. Their results showed some promise for better future models, with their model for classifying whether a user is indoors or outdoors being the most impressive. Successful models for predicting daily activities will certainly be important in a practical implementation of our system. This is because the body's response to alcohol consumption may share significant similarity to exercise, dance, or other activities.

Using smartphones and smartwatches, there is an active desire to create monitoring applications for serious health problems. Such as in \cite{Sharma:2014}, where Vinod Sharma, Kunal Mankodiya, Fernando De La Torre, \textit{et. al.}, developed a smartwatch-smartphone system for the monitoring and analysis of data from patients with Parkinson Disease. This system, named SPARK, includes the analysis of speech and detection of: facial tremors, dyskinesia, and freezing of gait. Their system is intended to provide useful recommendations to physicians based on the collected information. They concluded noting some potential problems of a full implementation of their system, the most relevant problem being misplacement of the sensors. This may also be a problem for us considering the potential importance of the motion-based accelerometer and gyroscope data.

### 2015-08-24 manuscript/content/system.tex
Switch format and other changes
Paper additions and cleanup.

\section{System Architecture}
A custom Android application (Figure \ref{app}) was designed to be used during the study. This application collected and locally stored sensor data from a Microsoft Band smartwatch in a CSV file. The following data was collected from the Microsoft Band:
\begin{itemize}
	\item Accelerometer
	\item Gyroscope
	\item Pace
	\item Speed
	\item Distance traveled
	\item Motion type (idle, walking, etc.)
	\item Heart rate
	\item Pedometer
	\item Skin temperature
	\item Ultraviolet level
	\item Contact (if smartwatch is on)
	\item Calories burned
\end{itemize}
Additional data continuously collected and stored in the CSV by the application included:
\begin{itemize}
	\item GPS coordinates from the smartphone 
	\item Button that was pressed whenever the user took a drink
	\item Input field used to periodically record the BAC measured by a separate breathalyzer.
\end{itemize}

\begin{figure}[H]
	\centering
	\includegraphics[scale=.20]{smartwatch.png}
	\caption{Capture of the Android application}
	\label{app}
\end{figure}

User information was also recorded and stored through the application before data collection began. Each user was given an unique user id (UUID) and the user profile included age, sex, body mass index, and blood type which was also stored locally as a JSON file.

After the data had been collected, the stored csv file and json file were uploaded to the web server and data inserted into mySQl database by a php script. Data stored on the MySQL database was then displayed on a web client by a geographical map.

\begin{figure}[H]
	\centering
	\includegraphics[scale=.35]{webpage.png}
	\caption{Current data visualization}
	\label{webpage}
\end{figure}

As of now, the web client displays the last sample for each user at the recorded location (Figure \ref{webpage}). Ideally in future systems, many samples will be averaged across regions (Figure \ref{concept}).

\begin{figure}[H]
	\centering
	\includegraphics[scale=.23]{webconcept.png}
	\caption{Ideal data visualization}
	\label{concept}
\end{figure}

The collected data was also used to train a machine learning model which would be implemented in a future Android application to predict user BAC or drunkenness.

\begin{figure}[H]
	\centering
	\includegraphics[scale=.30]{applicationsystem.png}
	\caption{Application system}
	\label{appsystem}
\end{figure}

Representation of this system can be found in Figure \ref{overallsystem}.

\begin{figure}[H]
	\centering
	\includegraphics[scale=.14]{overall_system.png}
	\caption{Overall system}
	\label{overallsystem}
\end{figure}

>>>>

\section{System Architecture}

% Present the overall system architecture here.

\subsection{Android-based Sensor Gateway}

% Describe the sensor gateway here.
